---
title: "Mallamaci HW4 Bank Feature Engineering"
author: "Anthony Mallamaci"
date: "2025-10-02"
output: html_document
---

Discretization regarding the age variable. I will use equal width binning where the continuous variable age will be split into 5 equally spaced bins.

```{r, echo=TRUE}
url <- "https://raw.githubusercontent.com/Antho028/MySTA551/refs/heads/main/Bank.Test.csv"
mydata <- read.csv(url)
#head(mydata)

mydata_binned <- mydata
mydata_binned$bin <- cut(
  mydata_binned$age,
  breaks = 5,
  labels = c("19-32", "33-46", "47-60", "61-74", "75-87")
)
#print(mydata_binned)

```


Variable encoding 

```{r, echo=TRUE}
url <- "https://raw.githubusercontent.com/Antho028/MySTA551/refs/heads/main/Bank.Test.csv"
mydata <- read.csv(url)
# head(mydata)

# Education level Varaible 
mydata$education <- factor(
  mydata$education,
  levels = c("unknown", "primary", "secondary", "tertiary"),
  ordered = TRUE
)

mydata$education_code <- as.numeric(mydata$education)

#head(mydata[, c("education", "education_code")])



# job variable
#install.packages("fastDummies")

# Load the package every time you start a new R session
library(fastDummies)

mydata <- fastDummies::dummy_cols(
  mydata,
  select_columns = "job",
  remove_first_dummy = FALSE   
)

# Check the new dummy columns
#head(mydata[, grep("^job_", names(mydata))])


# Variable contact 
mydata$contact <- trimws(tolower(mydata$contact))
#unique(mydata$contact)


mydata <- fastDummies::dummy_cols(
  mydata,
  select_columns = "contact",
  remove_first_dummy = FALSE   
)

#print(mydata[, grep("^contact_", names(mydata))])


# poutcome 

mydata <- fastDummies::dummy_cols(
  mydata,
  select_columns = "poutcome",
  remove_first_dummy = FALSE   
)

#print(mydata[, grep("^poutcome_", names(mydata))])



```


Transforming the balance variable to fix right skew. 

```{r, echo=TRUE}
library(MASS)

# Shift the variable so all values are positive
mydata$balance_shifted <- mydata$balance + 3314

# Fit a simple linear model 
bc_model <- lm(balance_shifted ~ 1, data = mydata)
any(mydata$balance_shifted < 0)
# Perform Box-Cox transformation to find optimal lambda
bc_result <- boxcox(bc_model, lambda = seq(-2, 2, by = 0.1))

# Extract the lambda with the maximum log-likelihood
lambda_opt <- bc_result$x[which.max(bc_result$y)]
cat("Optimal lambda:", lambda_opt, "\n")

# Apply the Box-Cox transformation using the optimal lambda
if (lambda_opt == 0) {
  mydata$balance_boxcox <- log(mydata$balance_shifted)
} else {
  mydata$balance_boxcox <- (mydata$balance_shifted^lambda_opt - 1) / lambda_opt
}

# Check the transformed variable
summary(mydata$balance_boxcox)
 

mydata$balance_log <- log(mydata$balance_shifted)

# Quick check: look at the distribution
hist(mydata$balance_log,
     main = "Histogram of Log-Transformed Balance",
     xlab = "log(balance_shifted)",
     breaks = 50,
     col = "skyblue")

qqnorm(mydata$balance_log)
qqline(mydata$balance_log, col = "red")

# Shapiro–Wilk normality test 
shapiro.test(sample(mydata$balance_log, min(5000, nrow(mydata))))

```

Using Box Cox to transform the balance variable didn't completely fix the problem but looking at the histogram it did make the balance variable closer to being normal. 